---
title: ""
output: html_document
date: '2022-03-25'
---

```{r setup, include=FALSE}
### essential libraries
`%$%` <- magrittr::`%$%`
library(tidyverse)

knitr::opts_chunk$set(echo = TRUE)
```

### Make Tables For MTP Documentation

In the Pediatric Molecular Open Targets high-level documentation <https://github.com/PediatricOpenTargets/documentation>, we have a table with the number of samples per cohort for the Datasets section.

**Data Retrieval:** In order to get the most up-to-date versions of the data tables, you need to clone the OpenPedCan-analysis GitHub repository <https://github.com/PediatricOpenTargets/OpenPedCan-analysis> and run the script `download-data.bash` to retrieve the latest data tables. There are more detailed instructions in the repository's README. The code in this document assumes that the OpenPedCan-analysis repository is downloaded locally and you would replace my local path `../../OpenPedCan-analysis` with your path to the repository.

<br><br>

### Cohort Sample Counts

We want to have a downloadable table with the number of subjects with each type of data in each cohort in each disease. Very simple counting and then cleanup of missing data.

---

Read in table. 

```{r}
vroom::vroom('../../OpenPedCan-analysis_FORK/data/v11/histologies.tsv') -> histologies
```

Count the number of subjects/participants per diease per sequencing type (aka experimental strategy) per cohort

```{r}
histologies %>%
# get unique combinations of patients, cohorts, types of data (experimental_strategy), 
# and diseases (cancer_group)
  distinct(Kids_First_Participant_ID, Kids_First_Biospecimen_ID, 
           cohort, experimental_strategy, cancer_group) %>% 
  count(cancer_group, cohort, experimental_strategy) %>% 
# sort for nice viewing on GitHub if people just want to look at the table
  arrange(cancer_group, cohort, experimental_strategy) %>%
# Clean up missing diseases; if the cohort is a disease cohort and cancer_group is NA
# that means the histology/disease hasn't been assigned and if it's GTEx or other
# healthy controls, there's no disease given in cancer_group because they're healthy
  mutate(cancer_group = case_when(cohort != 'GTEx' & is.na(cancer_group) ~ 'Undetermined',
                                  cohort == 'GTEx' & is.na(cancer_group) ~ 'Adult healthy controls',
                                  TRUE ~ cancer_group),
         cohort = ifelse(is.na(cohort) & experimental_strategy == 'Targeted Sequencing', 
                         'DGD', cohort)) %>%
# rename columns to be in line with the language used in the general MTP documentation
  select(disease = cancer_group, cohort, experimental_strategy,
         number_samples = n ) -> disease_subject_counts
```

Save one folder up since the `disease_subject_counts` table will be made available for download.

```{r}
# write_tsv(disease_subject_counts, '../disease_subject_counts.tsv')
```

Finally make a little table with the number of samples in each cohort/dataset for the dataset section. This was entered into Notion and then copy/pasted into the README

```{r}
histologies %>% 
  count(cohort)
```

<br><br>

